How environmental drivers of synchrony interact

### 2022-01-31 Paper/Paper.Rnw
various edits to Discussion and getting repo ready

%***DAN: Right now sup mat section S7 is not cited. It only has figures and tables, not sure I plan to 
%add anything to it other than that. Either get rid of it as a section, or else cite it in some of the places 
%where you currently cite the figures and tables in it. My current hunch is that the first of those options is
%better. Have to decide what to do with Fig. S1 in that case.

%Description of data after cleaning
The core of this paper is the new theory, but to help illustrate that theory, for goal G3 of the 
Introduction, we also apply
the theory to an exceptional dataset on giant kelp (\emph{Macrocyctis pyrifera}) dynamics off the CA coast, and 
associated environmental measurements. 
The data are based on a subset of those used by \cite{CastoraniSheppardKelp} and
\cite{Walter2021}, and data details are given in those papers and in SI section \ref{SM-sect:data}. We here
summarize the format of the data after a preparation and cleaning process was implemented. 
\cite{CastoraniSheppardKelp} and \cite{Walter2021} also provide an introduction to 
kelp ecology, and information on why giant kelp is an excellent species 
for studies of synchrony. After preparation, kelp data consisted of
$\Sexpr{length(SBlocstouse)+length(CC1locstouse)+length(CC2locstouse)}$ quarterly kelp abundance time series
from locations along the CA coast, each time series spanning from quarter $\Sexpr{quarters[1,3]}$ 
of $\Sexpr{quarters[1,2]}$ to quarter $\Sexpr{quarters[dim(quarters)[1],3]}$ of 
$\Sexpr{quarters[dim(quarters)[1],2]}$.   
Time series were grouped into three regions which were analyzed separately: a more northerly
central CA group of $\Sexpr{length(CC1locstouse)}$ locations (called CCA1); a more southerly central CA group of 
$\Sexpr{length(CC2locstouse)}$ locations (CCA2); and a group of $\Sexpr{length(SBlocstouse)}$ locations 
from southern CA, close to Santa Barbara
(called SB; see Fig. \ref{fig:map}). Each kelp measurement is an estimate of mean quarterly kelp biomass per unit 
useable habitat in a 500m stretch of coastline. We used coastline segments where kelp was persistent
through essentially all of the $\Sexpr{quarters[1,2]}$-$\Sexpr{quarters[dim(quarters)[1],2]}$ 
period (SI section \ref{SM-sect:data}). We also had estimates
of maximum wave height and mean nitrate concentration for each quarter and location. Both waves
and nutrients are known to influence kelp population dynamics and synchrony 
\citep{Cavanaugh2013,CastoraniSheppardKelp}.

@@

Data for the project are publicly archived \citep{SBCLTER2021}.
All computations were done in R version 3.4.4 on a laptop running Ubuntu Linux
16.04. Complete codes for the project workflow are at \url{https://github.com/reumandc/InteractingMoranEffects.git}.

@@

%Direct Moran effects
The way direct Moran effects work in our theory is fairly straightforward. 
For instance, the direct Moran effect
term in (\ref{rhoww_T1to3}) for nitrates, i.e., 
$|f_{\mcP^{(1)}}|^2 \rho_{\epsilon^{(1)} \epsilon^{(1)}}/|f_{\mcB}|^2$, 
equals the synchrony of the nitrate time series themselves, i.e.,
$\rho_{\epsilon^{(1)} \epsilon^{(1)}}$, times the 
timescale-specific strength of influence of nitrates on kelp, 
i.e., $|f_{\mcP^{(1)}}|^2$, and modulated
by the autoregressive nature of kelp dynamics,
$1/|f_{\mcB}|^2$. These components are plotted in Fig. \ref{fig:directMoran} (CCA1 and SB) and 
Fig. \ref{SM-fig:directMoran_SM_lag4} (CCA2) so their 
effects on synchrony can be assessed across timescales and regions.
Nitrates are henceforth identified with $\epsilon^{(1)}$ and waves with $\epsilon^{(2)}$. 
Both nitrates and waves fluctuate seasonally in CA (REFS). 
%***DAN: Fugure out a ref for this later
Thus nitrate and wave 
synchrony had a strong annual component (Fig. \ref{fig:directMoran}b,d,q,s), which produced 
some of the annual synchrony observed in kelp 
(Fig. \ref{fig:SummaryKelpFig}a,d, green and blue lines).
Nitrates and waves are also synchronous on long timescales $>8$y  
(Fig. \ref{fig:directMoran}l,n,aa,cc), possibly due to the North Pacific Gyre Oscillation 
[NPGO; \cite{CastoraniSheppardKelp}]. The long-timescale synchrony
in nitrates and waves produced some of the long-timescale
synchrony in kelp (Fig. \ref{fig:SummaryKelpFig}c,f, green and blue lines).
Kelp synchrony was stronger in CCA1 than in SB 
(Fig. \ref{fig:SummaryKelpFig}a-c v. d-f, black lines) in part because the synchrony of nitrates, and of 
waves, was more pronounced in CCA1 than in SB 
(Fig. \ref{fig:directMoran}b,g,l v. q,v,aa, and 
Fig. \ref{fig:directMoran}d,i,n v. s,x,cc), and also because waves
had a stronger influence on kelp in CCA1 than in SB (Fig. \ref{fig:directMoran}c,h,m v. r,w,bb).
Regression coefficients determining $f_{\mcP^{(1)}}$, $f_{\mcP^{(2)}}$
and $f_{\mcB}$ (Methods) are in Table \ref{SM-tab:regcoefs}.

%Interactions between Moran effects, CCA1
Interactions between Moran effects in CCA1 were parallel to one of the theoretical case studies, CaC1.
In CCA1, nitrates had a $1$q-delayed positive effect on observed 
kelp populations and waves had immediate negative effect \citep{Cavanaugh2011,Bell2015,Schiel2015}, 
just as $\epsilon^{(1)}$ effects in CaC 
were positive and delayed by $1$q and $\epsilon^{(2)}$ effects were negative and immediate.
Delayed effects are reflected, for CCA1, in Fig. \ref{fig:indirectMoran}a, and for 
CaC1, in Fig. \ref{fig:TheoryCaseA}j, both of which show 
that $f_{\mcP^{(1)}}$ has phase about $-\pi/2$ at the annual timescale. The delay is also 
reflected, for CCA1, in Table \ref{SM-tab:regcoefs}, where 
the entries for $p_0^{(1)}$ for CCA1 are close to $0$ and those for $p_1^{(1)}$ are positive.
Immediate negative effects of waves in CCA1 are reflected in Fig. \ref{fig:indirectMoran}b,g,l
and for CaC1 in Fig. \ref{fig:TheoryCaseA}j, both of which show 
that $f_{\mcP^{(2)}}$ has phase $-\pi$ at all timescales. See also the $p_0^{(2)}$
coefficients for CCA1 in Table \ref{SM-tab:regcoefs}, which are negative.
In CCA1, annual peak nitrates tend to come in spring, whereas annual peak waves 
tend to come in winter (REFS). 
%***DAN: add refs later
Thus annual nitrate peaks tend to lag annual 
wave peaks by $1$q, just as $\epsilon^{(1)}$ peaks lagged $\epsilon^{(2)}$
peaks by $1$q in CaC1.
This is reflected, for CCA1, in Fig. \ref{fig:indirectMoran}d, and for CaC1 
in Fig. \ref{fig:TheoryCaseA}k, both of which show that $\rho_{\epsilon^{(1)}\epsilon^{(2)}}$
has a phase of about $-\pi/2$ at the annual timescale.
Thus lags compound on the annual timescale, in the CCA1 case as in the theoretical case: the annual positive 
effects of nitrates on kelp tend to come in summer, and the annual negative effects of waves come in winter,
producing reinforcing interactions between Moran effects. Likewise, the $\pi/2$ phase 
of $f_{\mcP^{(1)}}\overline{f_{\mcP^{(2)}}}$ on annual timescales (Fig. \ref{fig:indirectMoran}c for CCA1, Fig. \ref{fig:TheoryCaseA}k for CaC1)
approximately cancels the $-\pi/2$ phase of $\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ on annual timescales
(Fig. \ref{fig:indirectMoran}d for CCA1, Fig. \ref{fig:TheoryCaseA}k for CaC1), so that  
$f_{\mcP^{(1)}}\overline{f_{\mcP^{(2)}}}\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ has phase 
approximately $0$, making the interaction-effect term 
$\Rp(f_{\mcP^{(1)}}\overline{f_{\mcP^{(2)}}}\rho_{\epsilon^{(1)}\epsilon^{(2)}})$ strongly positive 
on annual timescales for both the theoretical and emprical cases 
(Fig. \ref{fig:indirectMoran}e for CCA1, Fig. \ref{fig:TheoryCaseA}k for CaC1). 
Interactions between Moran effects on long-timescales ($>8$y) were also similar for both CCA1 and
CaC1, and for the same reason in both cases: on long timescales, sub-annual lags
are inconsequential, and the positive effects of nitrates/$\epsilon^{(1)}$ and the negative
effects of waves/$\epsilon^{(2)}$ therefore lead to negative interactions.
Interactions between Moran effects in CCA2 operated similarly to CCA1 (Fig. \ref{SM-fig:indirectMoran_SM_lag4}).

@@

%*DONE Summary
We provided a general mathematical theory of interactions between Moran effects. When two related spatially 
synchronous environmental drivers both influence a set of populations across a landscape, interactions can make 
synchrony in the populations either substantially stronger or markedly weaker 
than would be expected if the 
two drivers were unrelated, a new mechanism of population spatial synchrony.
Our general theory illuminates precisely how timings of 
influences of the Moran drivers on populations can interact with relationships between the
drivers to produce interaction effects, which may vary by timescale in both their strength and
sign. Moran effects are very common. Interactions between Moran effects have been tested for, and observed, twice previously, and may
be common [\cite{Sheppard2019,CastoraniSheppardKelp}; see also below]. 
Our study is the first to provide a comprehensive theory of the new mechanism. 
We used our theory and several case studies based on models and 
California kelp populations to provide improved intuition about the mechanism.

%*DONE arguments about likelihood that interactions will be common, extending earlier arguments
We argue that it is likely that interactions between Moran effect are common. Most species
are influenced by more than one environmental driver. Drivers are frequently spatially autocorrelated,
and are also often related to each other because of their common origin in underlying meteorological 
phenomena including major climatic oscillations such as 
the NPGO or El Ni\~{n}o Southern Oscillation (ENSO).
Driver pairs which may commonly produce interacting Moran effects include instances where the same quantity 
is measured in distinct parts of the year (e.g., spring and summer temperatures, or March and April rainfall,
or December and January snowfall); or when distinct variables are measured in the same part of the 
year (e.g., spring temperature and precipitation). Future work should systematically 
investigate cross synchrony of environmental variables, and could then use our theory
to directly assess the potential for interacting Moran effects to occur under a variety of biological scenarios. Since 
interactions between Moran effects have been discovered in both of the two systems so far 
tested \citep{Sheppard2019,CastoraniSheppardKelp}, future work should also, when
data are available, test directly for interactions.

%*Revist intuition
We revisit and extend the intuition behind interacting Moran effects, using the central CA kelp
example as a vehicle \citep{CastoraniSheppardKelp}. Large winter waves 
have immediate negative effects on kelp in central CA, whereas the positive
effects of spring nitrates manifest in summer. So nutrient and wave effects can 
reinforce each other in producing annual oscillations: large kelp increases in summer due 
to abundant nutrients can be followed by big crashes in winter due to waves, both factors combining
to accentuate the annual cycle. Thus positive
interactions between Moran effects on annual timescales occur whenever 
years with above-average waves coincide with years with plentiful nutrients in other locations:
if a large-wave year in location A coincides with a high-nutrient year in location B, 
both locations will tend to have bigger annual fluctuations that year, accentuating 
annual-timescale synchrony. Sub-annual lags and delays make essentially no difference, 
however, on long timescales. On long timescales, large-wave and abundant-nutrient years
counteract each other. If a multi-year period of large waves in location A coincides with
a multi-year period of abundant nutrients in location B, the multi-year-average kelp abundance
in A will tend to be reduced, whereas the multi-year-average kelp abundance in B will
tend to be augmented, reducing long-timescale synchrony. Lags and interactions between
drivers must always be compared to the timescale of interest to determine interaction effects,
as in Fig. \ref{fig:IntroTheory}, and as captured mathematically in (\ref{rhoww_T1to3}).

%*DONE Comparison to Castorani et al paper, and maybe also a bit on comparison to the Sheppard et al 2019 paper
Our results about kelp were consistent with those of \cite{CastoraniSheppardKelp}, though
the two papers have complementary emphases. In spite of numerous methodological choices which differed between the 
two studies, both our results and \cite{CastoraniSheppardKelp}
show positive effects of both nutrients and waves on synchrony on both annual and long ($>4$y)
timescales in both central and southern CA. And both studies
show positive interactions between nutrient and wave Moran effects on annual timescales
in central CA, but negative interactions in southern CA at annual timescales and in both central
and southern CA at long timescales. The Fourier approach of this study was designed to facilitate 
mathematical examination of interactions between Moran effect as a general mechanism, whereas the 
wavelet approach of \cite{CastoraniSheppardKelp} was instead optimized for detecting interactions 
and identifying Moran mechanisms in data, in spite of non-stationarity and
other complicating features which are present in many ecological datasets. The study of
\citep{Sheppard2019} developed the wavelet methods applied by \cite{CastoraniSheppardKelp}.
The approach of this study relates indirectly to the approach of \cite{Anderson2021}, though that
study concerned different research questions.

%*DONE relationship to earlier papers comparing synchrony of drivers and pops
It has been a frequent topic of research why populations are often less synchronous, or sychronous over a smaller
spatial extent, than might be expected given the strength or extent of 
synchrony of environmental drivers \citep{Herfindal2022}.
Antagonistic (negative) interaction effects between Moran drivers may be one common and previously
unrecognized mechanism contributing to this discrepancy. On the other hand,
a recent paper descibed an ``enhanced Moran effect'' by which specific 
patterns of temporal autocorrelation
in Moran drivers can theoretically cause greater synchrony in populations 
than in the drivers themselves \citep{Massie2015}.
Synergistic (positive) interaction effects is another mechanism by which populations can
be more synchronous than expected.

%*DONE How/why a frequency-specific approach is necessary for all this
An ever growing body of work shows that synchrony is, and should be analyzed as, a timescale-specific 
phenomenon \citep{Vasseur2007,Keitt2008,Vasseur2014,Sheppard2016,Defriez2016,Defriez2017,Defriez2017b,Desharnais2018,Anderson2021}.
The results of our present study extend this earlier work by showing that interactions between Moran effects 
are also timescale-specific, differing stongly in strength, and in sign, across timescales. 
A non-timescale-specific version of our theory could be obtained by appropriately integrating our main equation 
(\ref{rhoww_T1to3}) across frequencies. But doing so would reduce insight by conflating distinct and
often disparate processes. For instance, if the terms of (\ref{rhoww_T1to3}) were integrated 
across frequencies for our central CA kelp example, positive interactions between Moran drivers on short timescales 
(Fig. \ref{fig:SummaryKelpFig}a) would be summed with and partially cancel 
negative interactions on long timescales (Fig. \ref{fig:SummaryKelpFig}c),
falsely suggesting reduced importance of interaction effects. Our results also have implications given the 
short duration of most time series in ecology. 
Analyses based on short time series can only be expected to reveal Moran effects or interactions between Moran
effects occurring on short timescales, which our results show can differ substantially from long-timescale results. 
It has not escaped our notice that kelp time series even longer than ours, once 
available, may reveal Moran effects and interactions on timescales longer than ours which differ from
our results. In our view this provides another strong argument, among many others, in support of 
long-term monitoring of ecosystems.

%DECIDED NOT TO WRITE*Revisitation of diffs between SoCal and CentCal
%Note that we get differences in intereaction effects between central and southern CA because we 
%have lagged effects of nitrate on observed kelp density in central CA, but unlagged effects in southern
%CA (to see this, look at the regression coefficients computed to get the models used for the
%spectral analysis, which were saved). Careful inspection of my ms figures show this is the reason
%for the positive versus slightly negative interaction effects seen in central v. souther CA. 
%But I spke with Tom and Max and Kyle about this by email around 1/3/2022. They agreed that 
%probably it has something to do with the fact that kelp is ripped out by waves in central
%CA, so it takes a quarter to recover enough to be detectable from space (we can only see the 
%canopy from space, on the surface of the water). In southern CA, waves rip out kelp like that
%much less. So if kelp is always around and usually reaches the surface in southern CA, nitrate
%effects can be observed more quickly. There ought to be a Discussion paragraph on this, basiaclly
%saying things may be more complex that our analyses reveals with regard to some of the differences 
%we observed between central and southern CA. We didn't explore this because our goal was to illustrate 
%our theory, and our work does demonstrate the importance of interaction effects for the synchrony of kelp
%as observed from space. But future work could look at this more, certainly, perhaps using state-space models
%which can separate biological from observation processes.   
%   Tom also told me in a meeting zoom call that in central CA growth of new fronds will be slower because 
%there are not other fronds (left) on the same plant to provide a sugar subsidy to new fronds. In central CA,
%basically all fronds from the same plant are ripped out.

>>>>

%Description of data after cleaning
The core of this paper is the new theory, but to help illustrate that theory, for goal G3 of the 
Introduction, we also apply
the theory to an exceptional dataset on giant kelp (\emph{Macrocyctis pyrifera}) dynamics off the CA coast, and 
associated environmental measurements. 
The data are based on a subset of those used by \cite{CastoraniSheppardKelp} and
\cite{Walter2021}, and data details are given in those papers and in SI section \ref{SM-sect:data}. We here
summarize the format of the data after a preparation and cleaning process was implemented. 
\cite{CastoraniSheppardKelp} and \cite{Walter2021} also provide an introduction to 
kelp ecology, and information on why giant kelp is an excellent species 
for studies of synchrony. After preparation, kelp data consisted of
$\Sexpr{length(SBlocstouse)+length(CC1locstouse)+length(CC2locstouse)}$ quarterly kelp abundance time series
from locations along the CA coast, each time series spanning from quarter $\Sexpr{quarters[1,3]}$ 
of $\Sexpr{quarters[1,2]}$ to quarter $\Sexpr{quarters[dim(quarters)[1],3]}$ of 
$\Sexpr{quarters[dim(quarters)[1],2]}$.   
Time series were grouped into three regions which were analyzed separately: a more northerly
central CA group of $\Sexpr{length(CC1locstouse)}$ locations (called CCA1); a more southerly central CA group of 
$\Sexpr{length(CC2locstouse)}$ locations (CCA2); and a group of $\Sexpr{length(SBlocstouse)}$ locations 
from southern CA, close to Santa Barbara
(called SB; see Fig. \ref{fig:map}). Each kelp measurement is an estimate of 
mean quarterly kelp biomass (kg wet) per unit 
useable habitat ($\text{m}^2$) in a 500m stretch of coastline. We used coastline segments where kelp was persistent
through essentially all of the $\Sexpr{quarters[1,2]}$-$\Sexpr{quarters[dim(quarters)[1],2]}$ 
period (SI section \ref{SM-sect:data}). We also had estimates
of maximum wave height and mean nitrate concentration for each quarter and location. Both waves
and nutrients are known to influence kelp dynamics and synchrony 
\citep{Cavanaugh2013,CastoraniSheppardKelp}.

@@

Data for the project are publicly archived \citep{SBCLTER2021}.
All computations were done in R version 3.6.3 on a laptop running Ubuntu Linux
16.04. Complete codes for the project workflow are at \url{https://github.com/reumandc/InteractingMoranEffects.git}.

@@

%Direct Moran effects
The way direct Moran effects work in our theory is fairly straightforward. 
For instance, the direct Moran effect
term in (\ref{rhoww_T1to3}) for nitrates, i.e., 
$|f_{\mcP^{(1)}}|^2 \rho_{\epsilon^{(1)} \epsilon^{(1)}}/|f_{\mcB}|^2$, 
equals the synchrony of the nitrate time series themselves, i.e.,
$\rho_{\epsilon^{(1)} \epsilon^{(1)}}$, times the 
timescale-specific strength of influence of nitrates on kelp, 
i.e., $|f_{\mcP^{(1)}}|^2$, and modulated
by the autoregressive nature of kelp dynamics,
$1/|f_{\mcB}|^2$. These components are plotted in Fig. \ref{fig:directMoran} (CCA1 and SB) and 
Fig. \ref{SM-fig:directMoran_SM_lag4} (CCA2) so their 
effects on synchrony can be assessed across timescales and regions.
Nitrates are henceforth identified with $\epsilon^{(1)}$ and waves with $\epsilon^{(2)}$. 
Both nitrates and waves fluctuate seasonally in CA \citep{Schiel2015}. 
Thus nitrate and wave 
synchrony had a strong annual component (Fig. \ref{fig:directMoran}b,d,q,s), which produced 
some of the annual synchrony observed in kelp 
(Fig. \ref{fig:SummaryKelpFig}a,d, green and blue lines).
Nitrates and waves are also synchronous on long timescales $>8$y  
(Fig. \ref{fig:directMoran}l,n,aa,cc), possibly due to the North Pacific Gyre Oscillation 
[NPGO; \cite{CastoraniSheppardKelp}]. The long-timescale synchrony
in nitrates and waves produced some of the long-timescale
synchrony in kelp (Fig. \ref{fig:SummaryKelpFig}c,f, green and blue lines).
Kelp synchrony was stronger in CCA1 than in SB 
(Fig. \ref{fig:SummaryKelpFig}a-c v. d-f, black lines) in part because the synchrony of nitrates, and of 
waves, was more pronounced in CCA1 than in SB 
(Fig. \ref{fig:directMoran}b,g,l v. q,v,aa, and 
Fig. \ref{fig:directMoran}d,i,n v. s,x,cc), and also because waves
had a stronger influence on kelp in CCA1 than in SB (Fig. \ref{fig:directMoran}c,h,m v. r,w,bb).
Regression coefficients determining $f_{\mcP^{(1)}}$, $f_{\mcP^{(2)}}$
and $f_{\mcB}$ (Methods) are in Table \ref{SM-tab:regcoefs}.

%Interactions between Moran effects, CCA1
Interactions between Moran effects in CCA1 were parallel to one of the theoretical case studies, CaC1.
In CCA1, nitrates had a $1$q-delayed positive effect on observed 
kelp populations and waves had immediate negative effect \citep{Cavanaugh2011,Bell2015,Schiel2015}, 
just as $\epsilon^{(1)}$ effects in CaC 
were positive and delayed by $1$q and $\epsilon^{(2)}$ effects were negative and immediate.
Delayed effects are reflected, for CCA1, in Fig. \ref{fig:indirectMoran}a, and for 
CaC1, in Fig. \ref{fig:TheoryCaseA}j, both of which show 
that $f_{\mcP^{(1)}}$ has phase about $-\pi/2$ at the annual timescale. The delay is also 
reflected, for CCA1, in Table \ref{SM-tab:regcoefs}, where 
the entries for $p_0^{(1)}$ for CCA1 are close to $0$ and those for $p_1^{(1)}$ are positive.
Immediate negative effects of waves in CCA1 are reflected in Fig. \ref{fig:indirectMoran}b,g,l
and for CaC1 in Fig. \ref{fig:TheoryCaseA}j, both of which show 
that $f_{\mcP^{(2)}}$ has phase $-\pi$ at all timescales. See also the $p_0^{(2)}$
coefficients for CCA1 in Table \ref{SM-tab:regcoefs}, which are negative.
In CCA1, annual peak nitrates tend to come in spring, whereas annual peak waves 
tend to come in winter \citep{Schiel2015,Bell2015}. 
Thus annual nitrate peaks tend to lag annual 
wave peaks by $1$q, just as $\epsilon^{(1)}$ peaks lagged $\epsilon^{(2)}$
peaks by $1$q in CaC1.
This is reflected, for CCA1, in Fig. \ref{fig:indirectMoran}d, and for CaC1 
in Fig. \ref{fig:TheoryCaseA}k, both of which show that $\rho_{\epsilon^{(1)}\epsilon^{(2)}}$
has a phase of about $-\pi/2$ at the annual timescale.
Thus lags compound on the annual timescale, in the CCA1 case as in the theoretical case: the annual positive 
effects of nitrates on kelp tend to come in summer, and the annual negative effects of waves come in winter,
producing reinforcing interactions between Moran effects. Likewise, the $\pi/2$ phase 
of $f_{\mcP^{(1)}}\overline{f_{\mcP^{(2)}}}$ on annual timescales (Fig. \ref{fig:indirectMoran}c for CCA1, Fig. \ref{fig:TheoryCaseA}k for CaC1)
approximately cancels the $-\pi/2$ phase of $\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ on annual timescales
(Fig. \ref{fig:indirectMoran}d for CCA1, Fig. \ref{fig:TheoryCaseA}k for CaC1), so that  
$f_{\mcP^{(1)}}\overline{f_{\mcP^{(2)}}}\rho_{\epsilon^{(1)}\epsilon^{(2)}}$ has phase 
approximately $0$, making the interaction-effect term 
$\Rp(f_{\mcP^{(1)}}\overline{f_{\mcP^{(2)}}}\rho_{\epsilon^{(1)}\epsilon^{(2)}})$ strongly positive 
on annual timescales for both the theoretical and emprical cases 
(Fig. \ref{fig:indirectMoran}e for CCA1, Fig. \ref{fig:TheoryCaseA}k for CaC1). 
Interactions between Moran effects on long-timescales ($>8$y) were also similar for both CCA1 and
CaC1, and for the same reason in both cases: on long timescales, sub-annual lags
are inconsequential, and the positive effects of nitrates/$\epsilon^{(1)}$ and the negative
effects of waves/$\epsilon^{(2)}$ therefore lead to negative interactions.
Interactions between Moran effects in CCA2 operated similarly to CCA1 (Fig. \ref{SM-fig:indirectMoran_SM_lag4}).

@@

%*DONE Summary
We provided a general mathematical theory of the new mechanism of 
interactions between Moran effects. When two related spatially 
synchronous environmental drivers both influence a set of populations across a landscape, interactions can make 
synchrony in the populations either substantially stronger or markedly weaker 
than would otherwise be expected.
Our general theory illuminates precisely how timings of 
influences of the drivers on populations can interact with relationships between the
drivers to alter Moran effects. Interactions may vary by timescale in both their strength and
sign. We used our theory and several case studies based on models and 
kelp populations to provide intuition about the new mechanism (see also below).
Moran effects are ubiquitous. Interactions 
between Moran effects have been tested for only twice previously, and were observed
on both occassions; for this and other reasons (see below) interactions may
also be common \citep{Sheppard2019,CastoraniSheppardKelp}. 
Moran interactions are therefore
a newly recognized and potentially widespread aspect of a fundamental means (Moran effects) by which 
environmental factors influence populations
and diverse phenomena such as ecosystem stability \citep{Wilcox2017}, population cycling \citep{Anderson2021} 
and extinction \citep{Ghosh2020}.
Climate change is altering many aspects of environmental variables, including their means, variances and
spatial correlations \citep{Lyon2019,Keelings2020}, as well as 
relationships between environmental variables and the nature of 
their influences on populations. There is therefore potential for climate change to alter interactions 
between Moran effects, in ways our new theory may help researchers to understand (see below).  
The augmented fundamental understanding of Moran effects which we have provided may 
substantially benefit both basic \citep{Liebhold2004} 
and applied \citep{Hansen2020,Larsen2021,Herfindal2022} ecological research.

%*DONE arguments about likelihood that interactions will be common, extending earlier arguments
We argue that it is likely that interactions between Moran effect are common. Most species
are influenced by more than one environmental driver. Drivers are frequently spatially autocorrelated,
and are also often related to each other because of their common origin in underlying meteorological 
phenomena, including major climatic oscillations such as 
the NPGO or El Ni\~{n}o Southern Oscillation (ENSO).
Driver pairs which may commonly produce interacting Moran effects include instances where the same quantity 
is measured in distinct parts of the year (e.g., spring and summer temperatures, or March and April rainfall,
or December and January snowfall); or when distinct variables are measured in the same part of the 
year (e.g., spring temperature and precipitation). Future work should systematically 
investigate cross synchrony ($\rho_{\epsilon^{(1)} \epsilon^{(2)}}$) of environmental variables, and could then use our theory
to directly assess the potential for interacting Moran effects to occur under a variety of biological scenarios. Since 
interactions between Moran effects have been discovered in both of the two systems so far 
tested \citep{Sheppard2019,CastoraniSheppardKelp}, future work should also, when
sufficient data are available, test directly for interactions. Temperature and precipitation variables
measured in the same season may be particularly promising candidates for producing interactions because of
the well known joint influence of these variables on plants. 
\cite{Defriez2017b} tested for interactions between Moran influences
of temperature and precipitation on aspects of synchrony of a terrestrial plant biomass index, finding no 
evidence for interactions. But their data were global in extent and spatially coarse, their time series were 
short, and they used statistical methods more aligned with their goals of 
understanding the geography of synchrony than with our present goals of understanding 
synchrony itself. So temperature-precipitation interactions may still be found in future work
by applying our theory and the methods of \cite{Sheppard2019} and \cite{CastoraniSheppardKelp} to
suitable data. Likewise, rainfall variables measured in distinct portions
of the growing season may also be promising candidates for finding interaction effects, for similar reasons.

%*DONE Revist intuition
We revisit and extend the intuition behind interacting Moran effects, using the central CA kelp
example as a vehicle. Large winter waves 
have immediate negative effects on kelp in central CA, whereas the positive
effects of spring nitrates manifest in summer. So nutrient and wave effects can 
reinforce each other in producing annual oscillations: large kelp increases in summer due 
to abundant nutrients can be followed by big crashes in winter due to waves, both factors combining
to accentuate the annual cycle. Thus positive
interactions between Moran effects on annual timescales occur whenever 
years with above-average waves coincide with years with plentiful nutrients in other locations:
if a large-wave year in location A coincides with a high-nutrient year in location B, 
both locations will tend to have bigger annual fluctuations that year, accentuating 
annual-timescale synchrony between the locations \citep{CastoraniSheppardKelp}. 
Sub-annual lags and delays make essentially no difference, 
however, on long timescales. On long timescales, large-wave and abundant-nutrient years
counteract each other: if a multi-year period of large waves in location A coincides with
a multi-year period of abundant nutrients in location B, the multi-year-average kelp abundance
in A will tend to be reduced, whereas the multi-year-average kelp abundance in B will
tend to be augmented, reducing long-timescale synchrony. Lags and interactions between
drivers must always be compared to the timescale of interest to determine interaction effects,
as in Fig. \ref{fig:IntroTheory}, and as captured rigorously in our theory.

%*DONE Comparison to Castorani et al paper, and maybe also a bit on comparison to the Sheppard et al 2019 paper
Our results about kelp were consistent with those of \cite{CastoraniSheppardKelp}, though
these two studies use distinct methods; the studies complement each other in important ways. In spite of 
numerous methodological choices which differed between the 
two studies, both our results and \cite{CastoraniSheppardKelp}
show positive effects of both nutrients and waves on synchrony on both annual and long ($>4$y)
timescales in both central and southern CA. And both studies
show positive interactions between nutrient and wave Moran effects on annual timescales
in central CA, but negative interactions in southern CA on annual timescales and in both central
and southern CA on long timescales. The Fourier approach of our study was designed to facilitate 
mathematical examination of interactions between Moran effect as a general mechanism and development
of general theory, whereas the 
wavelet approach of \cite{CastoraniSheppardKelp} was instead optimized for detecting interactions 
and identifying Moran mechanisms in data, in spite of non-stationarity and
other complicating features which are present in many ecological datasets. The study of
\citep{Sheppard2019} developed the wavelet methods applied by \cite{CastoraniSheppardKelp}.
The modelling approach of this study relates indirectly to the approach of \cite{Anderson2021}, though that
study concerned different research questions.

%*DONE relationship to earlier papers comparing synchrony of drivers and pops
It has been a frequent topic of research why populations are often less synchronous, or sychronous over a smaller
spatial extent, than might be expected given the strength and extent of 
synchrony of environmental drivers \citep{Herfindal2022}.
Antagonistic interaction effects between Moran drivers may be one common and previously
unrecognized mechanism contributing to this discrepancy. On the other hand,
two recent papers descibed ``enhanced Moran effects'' by which specific 
patterns of temporal autocorrelation
in Moran drivers can theoretically cause greater synchrony in populations 
than in the drivers themselves \citep{Massie2015,Desharnais2018}.
Synergistic interaction effects is another mechanism by which populations can
be more synchronous than expected.

%*DONE Revisit the kendall paper, briefly. Talk about possible three-way interactions.
\cite{Kendall2000} considered interactions between mechanisms of synchrony. Their research 
questions were distinct from ours, but combining their viewpoints and ours may lead to valuable
future work. \cite{Kendall2000} considered interactions between dispersal and Moran causes of 
synchrony, whereas we analyzed interactions between two Moran drivers, a distinct context.
But it seems likely that numuerous ecological systems involve important dispersal effects on 
synchrony as well as influences of at least two Moran effects. It may be necessary in future
work to consider three-way interactions between dispersal and two Moran drivers and their 
influence on synchrony. Dispersal can readily be added to our modelling framework.
\cite{Desharnais2018} performed a spectral analysis on a model similar to ours
which included dispersal but which only considered two habitat patches and one environmental driver.

%*DONE Climate change
Climate change has the potential to influence interactions between Moran effects in two specific ways
which can be illuminated by our theory, and this potential should be investigated in
future work. Examining the third term of (\ref{rhoww_T1to3}), climate change could
alter interaction effects if it: 1) alters the term $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
quantifying cross synchrony between Moran drivers; or 2) alters one of the terms
$f_{\mcP^{(1)}}$ or $f_{\mcP^{(2)}}$ specifying the nature of the influence of the 
environmental variables $\epsilon^{(1)}$ and $\epsilon^{(2)}$, respectively, on populations. As advocated above,
the term $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ should be systematically computed in future work, for a variety 
of environmental variables, to assess whether interactions between Moran effects are likely to
be general. As part of that process, the potential for changes in $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
could also be assessed, by using either time-windowed versions or wavelet adaptations 
of this statistic, applied to either long-term climate records or future climate scanarios. 
Differences in $f_{\mcP^{(1)}}$ were responsible for differences
in the nature of Moran interactions between central CA and southern CA (Results), specifically
differences between the two regions in the lag of nitrate effects on kelp populations. 
If climate change modifies environmental effects on populations in a related way it would 
be expected to produce similarly large changes in Moran interactions. Nitrate effects on kelp 
are probably delayed by a quarter in central CA because winter waves commonly remove kelp,
and it takes time for kelp to grow to the surface and become detectable by satellite, only
then revealing the effects of elevated nitrates. In southern CA, kelp are less likely to be
removed by winter waves. Effects of elevated nitrates on the growth of kelp stands already reaching
the surface can be immediately observed. Climate change may similarly alter 
lags and delays associated with environmental effects on populations in at least two ways:
1) by altering species phenology; and 2) by increasing or decreasing growth rates and thereby
decreasing or increasing delays (respectively). Though it is too early to conclude that
effects on Moran interactions are among the important impacts of climate change, we feel the mechanisms
outlined above are sufficient to warrant further investigation.
